From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-28 paper/paper.Rmd
refs, sp

To the empricical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of stastical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuciance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transistion from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empricial thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empricial communities by 
seeing noise not as mathematical curiousity or statistical bugbear, but as a
source for new opportunities for inference.

The gap between theory-focused specialists and more empirically minded non-specialists
is particularly acute in the ecological literature on stochastic processes, making
a review of

As the literature on stochastic modeling in ecology is both extensive and often
technical, and so rather than attempt a comprehensive review, this synthesis
will focus on the simplest examples necessary to illustrate this broader trend
in the changing way we think about the role of noise.
To that end, I will focus on examples from simple population models, echoing
the motivation of @Bartlett1960 to focus on basic theoretical aspects common
across models. In so doing, we will ignore the equally rich literature of 
stochastic models in evolution [@Lenormand2009] and population genetics
[e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
readers, I will rely on the simplest examples necessary, with only passing
mention of the rich literature extending these concepts to more realistic
models.  It is easy to critique this choice: indeed, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticity literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
sacrificing realism for generality.

In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  Appendix B provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

What is noise? Where does it come from? 
Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level in deterministic settings. The notion of a deterministic
"skeleton" -- responsible for the phenomena of interest -- that is obscured
from view by the nusiance of noise has slowly given way to recognition
that stochastic and deterministic elements not only interact in 
interesting ways [@Coulson2004] but also share a common origin [e.g. @Black2012].
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise.

Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
This literature has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intriniscally 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
"population size" or "population density" is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

It is often suggested that stochasticity is "inherent" or "fundamental,"
e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinisic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

@@

Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e. Gaussian white noise process) for demographic and environmental stochasticity respectively.  When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g. @Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements. Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.

@@

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level.  
The algorithm is merely a simple and direct implemtentation of the master
equation, progressing in random step sizes deterimed
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free[^2] stochastic ecological models.

[^2]: that is, free from the approximation made by SDE models as we see in the
van Kampen example.  All models are, of course, only approixmations.

As our objective is to tie the origins of noise more closely to biological
processes, it will be helpful to make the notion of a master equation concrete
with a spefiic example.  We will focus on the classic case of @Levins1969 patch model,
to illustrate the Gillespie algorithm and the van Kampen system size expansion

@@

We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convience but leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formuluations.

@@

van Kampen's approximation follows a central-limit theorem style argument established rigorously in
the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Thereom 4.4. @Kurtz1978].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctionations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

@@

This result is striking for a variety of reasons.  First, we see the recovery of the 
"deterministic skeleton," in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of varation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and nonlinearity in noise-driven phenomena such as stochastic
transitions between alternate stable states.

@@

$$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
which will also prove useful later on.  Note this expression simply states that the equilbrium
variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
the fluctuation-disspation theorem) and that correlations fall off exponentially at a rate given by this slope.   
We will see both autocorrelation and variance play similar roles to infer this eigenvalue when we 
reach the role of noise as informer.

In the interests of a theoretical synthesis, I observe that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
or the review in @Ovaskainen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonomous with the diffusion approximation.
In brief, the diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskainen2010)

@@

This better-known approach has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological intepretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskainen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.

@@

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly autocorrelated noise 
(see derivation, Appendix B).  Auto-correlated or "colored" environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of autocorrelation 
in environmental time series rather than a process-based derivation.
As the expansion clearly shows, any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation in Appendix B makes clear,
it is rather the the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the autocorrelation time of the environmental dynamics and population dynamics.)
An environment with comparatively short autocorrelation scale will act
effectively as white noise upon the population dynamics.

The van Kampen expansion for extrinsic factors (appendix B) also highlights
the importance of how environmental noise couples to the population dynamics.
A mechanistic description of environmental noise acts directly on parameters
of the ecological model: for instance, in our Levins' model we may imagine
that the extinction rate of patches does not occur at some fixed rate $e$
but rather at some stochastically fluctuating rate of a given mean, variance,
and autocorellation.  The system size expansion illustrates how we replace
a model where stochasticity is nested inside some parameter into the simpler
format of the canonical equation where this contribution can be merely added
on as Gaussian noise at the end of a deterministic skeleton.  (The expansion
derives that the environmental noise is proportional to the square of the derivative
of the deterministic skeleton with respect to the environmental variable,
see appendix B.) Thus for the Levin's model, we can capture the fluctuations
introduced by demographic and environmental noise at equilibrium as
a Gaussian distribution with variance given by the sum of of demographic
and environmental contributions:

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise. Unsurprisingly few topics have
recieved greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing autocorrelated noise [e.g. @Ripa; @Ripa1999; 
@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.

>>>>

To the empirical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of statistical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuisance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transition from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empirical thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empirical communities by 
seeing noise not as mathematical curiosity or statistical bugbear, but as a
source for new opportunities for inference.

The gap between theory-focused specialists and more empirically minded non-specialists
is particularly acute in the ecological literature on stochastic processes, making
a review of 
<!-- FIXME -->

As the literature on stochastic modeling in ecology is both extensive and often
technical, and so rather than attempt a comprehensive review, this synthesis
will focus on the simplest examples necessary to illustrate this broader trend
in the changing way we think about the role of noise.
To that end, I will focus on examples from simple population models, echoing
the motivation of @Bartlett1960 to focus on basic theoretical aspects common
across models. In so doing, we will ignore the equally rich literature of 
stochastic models in evolution [@Lenormand2009] and population genetics
[e.g. @Ewens2004; @Gillespie2004]. In an effort to make these themes accessible to more empirical
readers, I will rely on the simplest examples necessary, with only passing
mention of the rich literature extending these concepts to more realistic
models.  It is easy to critique this choice: indeed, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticity literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
sacrificing realism for generality.

In this spirit, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  Appendix B provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

What is noise? Where does it come from? 
Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level in deterministic settings. The notion of a deterministic
"skeleton" -- responsible for the phenomena of interest -- that is obscured
from view by the nuisance of noise has slowly given way to recognition
that stochastic and deterministic elements not only interact in 
interesting ways [@Coulson2004] but also share a common origin [e.g. @Black2012].
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise.

Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
This literature has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intrinsically 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
"population size" or "population density" is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extrinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

It is often suggested that stochasticity is "inherent" or "fundamental,"
e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinsic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

@@

Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e. Gaussian white noise process) for demographic and environmental stochasticity respectively.  When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g. @Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements. Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of auto-correlated, rather than than white noise of $\textrm{d}B_t^{(e)}$.

@@

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level.  
The algorithm is merely a simple and direct implementation of the master
equation, progressing in random step sizes determined
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free[^2] stochastic ecological models.

[^2]: that is, free from the approximation made by SDE models as we see in the
van Kampen example.  All models are, of course, only approximations.

As our objective is to tie the origins of noise more closely to biological
processes, it will be helpful to make the notion of a master equation concrete
with a specific example.  We will focus on the classic case of @Levins1969 patch model,
to illustrate the Gillespie algorithm and the van Kampen system size expansion

@@

We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convenience but leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.

@@

van Kampen's approximation follows a central-limit theorem style argument established rigorously in
the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Theorem 4.4. @Kurtz1978].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctuations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

@@

This result is striking for a variety of reasons.  First, we see the recovery of the 
"deterministic skeleton," in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an ensemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of variation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and non-linearity in noise-driven phenomena such as stochastic
transitions between alternate stable states.

@@

$$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
which will also prove useful later on.  Note this expression simply states that the equilibrium
variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
the fluctuation-dissipation theorem) and that correlations fall off exponentially at a rate given by this slope.   
We will see both auto-correlation and variance play similar roles to infer this eigenvalue when we 
reach the role of noise as informer.

In the interests of a theoretical synthesis, I observe that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
or the review in @Ovaskainen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
In brief, the diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskainen2010)

@@

This better-known approach has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological interpretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskainen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.

@@

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly auto-correlated noise 
(see derivation, Appendix B).  Auto-correlated or "colored" environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of auto-correlation 
in environmental time series rather than a process-based derivation.
As the expansion clearly shows, any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation in Appendix B makes clear,
it is rather the the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the auto-correlation time of the environmental dynamics and population dynamics.)
An environment with comparatively short auto-correlation scale will act
effectively as white noise upon the population dynamics.

The van Kampen expansion for extrinsic factors (appendix B) also highlights
the importance of how environmental noise couples to the population dynamics.
A mechanistic description of environmental noise acts directly on parameters
of the ecological model: for instance, in our Levins' model we may imagine
that the extinction rate of patches does not occur at some fixed rate $e$
but rather at some stochastically fluctuating rate of a given mean, variance,
and auto-correlation.  The system size expansion illustrates how we replace
a model where stochasticity is nested inside some parameter into the simpler
format of the canonical equation where this contribution can be merely added
on as Gaussian noise at the end of a deterministic skeleton.  (The expansion
derives that the environmental noise is proportional to the square of the derivative
of the deterministic skeleton with respect to the environmental variable,
see appendix B.) Thus for the Levins' model, we can capture the fluctuations
introduced by demographic and environmental noise at equilibrium as
a Gaussian distribution with variance given by the sum of of demographic
and environmental contributions:

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise. Unsurprisingly few topics have
received greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.
